\let\clearforchapter\par % cheating, but saves some space

\chapter{Introduction}

Scout is an experimental programming language that combines
sequential, task- and data-parallel computing constructs along with a
set of specialized constructs for the visualization and analysis of
data. In addition, the language exposes a set of \emph{first-class}
data types that represent a variety of computational meshes.  These
features are provided via extensions to the C programming language.
In this manual we assume the reader is familiar with some of the basic
ideas behind parallel programming and fully familiar with the C
language.  This chaper provides a high-level introduction to the
compiler, data types, and language constructs -- more complete details
of each of these topics is provided in the remaining chapters.

\section{Scout and the C Programming Language}
\label{ch1:scout-clang}

Scout extends the C programming language to support a new set of
abstractions and domain-specific constructs.  Some of the main
domain-specific constructs in Scout are explicit mesh declarations,
instantiations and parallel computation over the various components of
the mesh (e.g. cells, vertices, edges) or over array elements.  Scout
builds upon the fundamental types of C to include higher-level
abstractions -- such as computational meshes and the associated fields
stored on a mesh.  From this perspective, Scout is a higher-level
language than C.

Since Scout is based on the open-source Clang and LLVM Compiler Infrastructure, all of 
the C language features are available. 
Scout extends the fundamental data types
available in C to include vector types of two, three, and four components.  Scout's 
syntax for vector types follows closely with those used by the OpenGL Shading Language
and NVIDIA's C for CUDA.  Section~\ref{ch:datatypes} below covers these topics in more 
detail.

\section{Getting Started}
\label{ch1:started}

The best way to learn any programming language is to get your hands dirty.  The classic
{hello world} program does not need to rewritten for Scout, since Scout fully supports
C.  Therefore, as an introductory program to Scout, Listing~\ref{lst-forallmesh}  
illustrates a simple one-dimensional mesh transformation.  It uses the \texttt{mesh}
abstraction, the mesh's associated fields and the parallel \texttt{forall} construct.

\par\bigskip
\begin{lstlisting}[label=lst-forallmesh,caption={Simple mesh example.}]
uniform mesh MyMesh{
  cells:
    float a;
    float b;
    float sum;
};

int main(int argc, char** argv){

  MyMesh m[512];

  // initialization of a and b not shown

  forall cells c in m {
    sum = a + b;
  }

  return 0;
}
\end{lstlisting}
\par\bigskip\noindent

\section{Compiling a Scout Program}
\label{ch1:compiling}

Scout's compiler follows the characteristics of traditional (Unix) command line interfaces.  In 
the following example, the simple mesh program from Listing~\ref{lst-forallmesh} is compiled and 
executed from the command prompt:

\par\bigskip
\fcolorbox{scoutblue}{lightgray}{
  \parbox{\textwidth-10pt}{%
    \texttt{\small \$ scc simplemesh.sc} \cmdcmt{compiler produces default executable ''a.out''...}\\      
    \texttt{\small \$ ./a.out}\\
    \texttt{\small \$ \_ }
  }
}
\par\bigskip\noindent

The previous compilation of the Scout program creates an executable that is targeted to run
sequentially on a CPU.  

%Scout can also be enabled to run CPU multithreading via the \texttt{-mt}
%option to \texttt{scc} or run on an NVIDIA GPU via the \texttt{-gpu} option to \texttt{scc}.

Scout can be enabled to run on an NVIDIA GPU via the \texttt{-gpu} option to \texttt{scc}.

% Do we have an scc man page anywhere?
%See the on-line \texttt{scc} man page, or Appendix~\ref{app-scc} for more information on the
%available command line options.

See the output to \texttt{scc -help} or Appendix~\ref{app-scc} for more information on the
available command line options.
